Elaboration with First-Class Implicit Function Types

### 2020-07-08 paper.tex
camery-ready update

\subsubsection{Note on terminology}

@@

Function types only differ from each other in notation: implicit
functions have the same rules as ``explicit'' functions. The primary purpose of
implicit function types is to \emph{guide elaboration}: the elaborator will at
times compute a type and branch on whether it is an implicit function.

We avoid a fully formal presentation in this paper. For examples of what such a
presentation looks like, see \cite{decidableconv} or \cite{coqnbe}. Some notes
on what is elided:
\begin{itemize}
  \item We use nameful notation and implicit weakening, i.e.\ whenever a term is
    well-formed in some context, it is assumed to be well-formed (as it is) in
    extended contexts. We also assume that any specifically mentioned name is
    fresh, e.g.\ when we write $\Theta,\,\alpha : A$, we assume that $\alpha$ is
    fresh in $\Theta$.  Formally, we would use de Bruijn indices for variables,
    and define variable renaming and parallel substitution by recursion on
    presyntax, e.g.\ as in \cite{autosubst}.
  \item Fixing any $\Theta$ metacontext, parallel substitutions of bound and
    defined variables form morphisms of a category, where the identity
    substitution $\id$ maps each variable to itself and composition
    $\blank\circ\blank$ is given by pointwise substitution. The action of
    parallel substitution on terms is functorial, i.e.\ $t[\sigma][\delta]
    \equiv t[\sigma\circ\delta]$ and $t[\id] \equiv t$, and typing is stable
    under substitution.
  \item
    Definitional equality is understood to be a congruence and an equivalence relation,
    which is respected by substitution and typing.
  \item
    We elide a number of well-formedness assumptions in rules. For instance, whenever
    a context appears in a rule, it is assumed to be well-formed. Likewise, whenever
    we have $\Theta|\Gamma\vdash t : A$, we assume that $\Theta|\Gamma\vdash A : \U$.
\end{itemize}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have helper functions
for implicit argument insertion, which can be used as a post-processing step
after inference. First, about the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification. Patterns are matched in top-down order.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.
\begin{mathpar}
\inferrule*[lab=insertion,flushleft]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert'(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           \\
            \einsert(\Theta,\,\theta,\,t,\,A)\hspace{0.3em}\in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           }
\end{mathpar}
\begin{alignat*}{3}
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert'\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert'\,\fail && :\equiv && \fail\\
  &\einsert\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A) && :\equiv&& \ereturn\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A)\\
  &\einsert\,(\Theta,\,\theta,\,t,\,A) && :\equiv&& \einsert'\,(\Theta,\,\theta,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail\\
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

\inferrule*[lab=infer]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
           {\einfer{t}{\Theta}{\Gamma}\,\in
             \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A)\}\cup\{\fail\}
           }
\end{mathpar}

@@

\subsubsection{Implicit argument insertion}

@@

\subsubsection{Starting and finishing elaboration} Given a surface term $t$,
we initiate elaboration by computing $\einfer{t}{\emptycon}{\emptycon}$. If this
succeeds, we get a $(\Theta,\,\theta,t')$ result. Elaboration is successful overall
if $\Theta = \emptycon$ and $\theta = \id$, i.e.\ no unsolved metas remain.

\subsubsection{Properties of elaboration}
First, elaboration is \emph{sound} in the sense that it only produces well-formed
output.

@@

\subsubsection{Omitted features}

@@

\subsubsection{On metatheoretical properties of the core syntax}
We naturally want to make sure that the extended core syntax is still sensible.
In relation to type theories, the following three properties are commonly
considered.
\begin{itemize}
  \item A theory is \emph{consistent} if not every closed type has a closed
    inhabitant. An inconsistent theory cannot be used as a \emph{logic}, from
    the propositions-as-types perspective \cite[Chapter~1]{hottbook}, because
    there every proposition is provable, so proofs do not distinguish truth from
    falsehood.
  \item A theory has the \emph{canonicity} property, if every closed term is
    definitionally equal to a canonical term. For example, every closed natural
    number is a numeral, every $\mathsf{Bool}$ is $\true$ or $\false$ and so
    on. A canonical theory is useful as a \emph{total programming language}; if
    canonicity fails, then some closed programs diverge or compute to stuck
    terms which cannot be viewed as values. In non-total languages, progress and
    type preservation constitute a weaker analogue of canonicity
    \cite[Chapter~6]{harper2016practical}.
  \item A theory has \emph{decidable conversion}, if definitional equality is
    decidable. Undecidable conversion usually implies undecidable type checking
    for type theories. Often, decidable conversion follows from a normalization
    procedure for open terms: we first normalize, then decide equality of normal
    forms.
\end{itemize}

@@

\subsubsection{Constancy constraints}
We use these constraints to get rid of curried function types as soon as we
learn that they are non-dependent. They are constraints in the usual sense in
unification algorithms (e.g.\ as in \cite{abel2011higher} or
\cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
new kind of context extension for metacontexts. The rules are given in Figure
\ref{fig:constancy}.

@@

\subsubsection{Algorithmic implementation of constraint solving}
\label{sec:algorithmic}
The above specification for constancy constraints is compact but not
particularly algorithmic: we just magically get new definitional equalities
whenever we have constraints in contexts. In our prototype implementation, we
implement eager removal of solvable constraints.

>>>>

\subsection{Note on Terminology}

@@

Function types only differ from each other in notation: implicit
functions have the same rules as ``explicit'' functions. The primary purpose of
implicit function types is to \emph{guide elaboration}: the elaborator will at
times compute a type and branch on whether it is an implicit function type.

We avoid a fully formal presentation in this paper. For examples of what such a
presentation looks like, see \cite{decidableconv} or \cite{coqnbe}. Some notes
on what is elided:
\begin{itemize}
  \item We use nameful notation and implicit weakening, i.e.\ whenever a term is
    well-formed in some context, it is assumed to be well-formed (as it is) in
    extended contexts. We also assume that any specifically mentioned name is
    fresh, e.g.\ when we write $\Theta,\,\alpha : A$, we assume that $\alpha$ is
    fresh in $\Theta$.  Formally, we would use de Bruijn indices for variables,
    and define variable renaming and parallel substitution by recursion on
    presyntax, e.g.\ as in \cite{autosubst}.
  \item Fixing any $\Theta$ metacontext, parallel substitutions of bound and
    defined variables form morphisms of a category, where the identity
    substitution $\id$ maps each variable to itself and composition
    $\blank\circ\blank$ is given by pointwise substitution. The action of
    parallel substitution on terms is functorial, i.e.\ $t[\sigma\circ\delta]
    \equiv t[\sigma][\delta]$ and $t[\id] \equiv t$, and typing is stable under
    substitution.
  \item
    Definitional equality is understood to be a congruence and an equivalence relation,
    which is respected by substitution and typing.
  \item
    We elide a number of well-formedness assumptions in rules. For instance, whenever
    a context appears in a rule, it is assumed to be well-formed. Likewise, whenever
    we have $\Theta|\Gamma\vdash t : A$, we assume that $\Theta|\Gamma\vdash A : \U$.
\end{itemize}

@@

Elaboration consists of two (partial) functions, checking and inferring, which
are defined by mutual induction on surface syntax. We also have helper functions
for implicit argument insertion, which can be used as a post-processing step
after inference. First, about the used notations:
\begin{itemize}
  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
    failure via $\fail$.
  \item We use pattern matching notation on core terms; e.g.\ we may match on
    whether a type is a function type. This assumes an evaluation/normalization
    procedure on core terms; but note that we already assume this in
    unification. Patterns are matched in top-down order.
  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$ and $\theta_1
    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
    cases. We do this to reduce the visual noise caused by threading composed
    metasubstitutions everywhere in the elaboration algorithm.
\end{itemize}
We present the specifications and definitions below, then we describe them in order.

\begin{mathpar}
\inferrule*[lab=insertion,flushleft]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
           {\einsert'(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           \\
            \einsert(\Theta,\,\theta,\,t,\,A)\hspace{0.3em}\in \{(\Theta,\,\theta,\,t,\,A)\,|\,
             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}
           }
\end{mathpar}
\begin{alignat*}{3}
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
  & && &&\sin\,\einsert'\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
  &\einsert'\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
  &\einsert'\,\fail && :\equiv && \fail\\
  &\einsert\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A) && :\equiv&& \ereturn\,(\Theta,\,\theta,\,(\lambda\{x\}.t),\,A)\\
  &\einsert\,(\Theta,\,\theta,\,t,\,A) && :\equiv&& \einsert'\,(\Theta,\,\theta,\,t,\,A)\\
  &\einsert\,\fail && :\equiv && \fail\\
\end{alignat*}
\begingroup
\allowdisplaybreaks
\begin{mathpar}
\inferrule*[lab=check]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
           {\echeck{t}{\Theta}{\Gamma}{A} \in
             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
           }

\inferrule*[lab=infer]
           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
           {\einfer{t}{\Theta}{\Gamma}\,\in
             \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
               (\Theta'|\Gamma[\theta]\vdash t' : A)\}\cup\{\fail\}
           }
\end{mathpar}

@@

\subsubsection{Implicit Argument Insertion}

@@

\subsubsection{Starting and Finishing Elaboration} Given a surface term $t$,
we initiate elaboration by computing $\einfer{t}{\emptycon}{\emptycon}$. If this
succeeds, we get a $(\Theta,\,\theta,t')$ result. Elaboration is successful overall
if $\Theta = \emptycon$ and $\theta = \id$, i.e.\ no unsolved metas remain.

\subsubsection{Properties of Elaboration}
First, elaboration is \emph{sound} in the sense that it only produces well-formed
output.

@@

\subsubsection{Omitted Features}

@@

We naturally want to make sure that the extended core syntax is still sensible.
In relation to type theories, the following three metatheoretical properties are
commonly considered.
\begin{itemize}
  \item A theory is \emph{consistent} if not every closed type has a closed
    inhabitant. An inconsistent theory cannot be used as a \emph{logic}, from
    the propositions-as-types perspective \cite[Chapter~1]{hottbook}, because
    there every proposition is provable, so proofs do not distinguish truth from
    falsehood.
  \item A theory has the \emph{canonicity} property, if every closed term is
    definitionally equal to a canonical term. For example, every closed natural
    number is a numeral, every $\mathsf{Bool}$ is $\true$ or $\false$ and so
    on. A canonical theory is useful as a \emph{total programming language}; if
    canonicity fails, then some closed programs diverge or compute to stuck
    terms which cannot be viewed as values. In non-total languages, progress and
    type preservation constitute a weaker analogue of canonicity
    \cite[Chapter~6]{harper2016practical}.
  \item A theory has \emph{decidable conversion}, if definitional equality is
    decidable. Undecidable conversion usually implies undecidable type checking
    for type theories. Often, decidable conversion follows from a normalization
    procedure for open terms: we first normalize, then decide equality of normal
    forms.
\end{itemize}

@@

\subsubsection{Constancy Constraints}
We use these constraints to get rid of curried function types as soon as we
learn that they are non-dependent. They are constraints in the usual sense in
unification algorithms (e.g.\ as in \cite{abel2011higher} or
\cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
new kind of context extension for metacontexts. The rules are given in Figure
\ref{fig:constancy}.

@@

\subsubsection{Algorithmic Implementation of Constraint Solving}
\label{sec:algorithmic}
The above specification for constancy constraints is compact but not
particularly algorithmic: we just magically get new definitional equalities
whenever we have constraints in contexts. In our prototype implementation, we
implement eager removal of solvable constraints.
